---
title: "Sampling subtype reference sequences from subtrees of a phylogenetic tree"
author: "Soo-Yon Rhee"
date: "September 30, 2017"
output: html_document
---
## Overviews

This R script has been described in the manuscript by Rhee, S-Y and Shafer, RW, "Geographically-Stratified Representative HIV-1 Group M pol Subtype and Circulating Recombinant Form Sequences". For each fasta sequence file in a given directory, this script performs the followings: (1) calculates TN93 pairwise distances (PWDs);  (2) creates a neighbor joining tree with the PWD matrix and roots it at the midpoint; (3) traverses the tree from the root to leaves and partitioned it into subtrees having a local median PWD (MPWD) less than a cutoff; (4) then selects one sequence per country from each subtree. In the manuscript, 47% (0.47) of the ranked PWDs was used to define the maximum MPWD for subtrees.

#### Usage

To run the script using the default parameters

Rscript -e "rmarkdown::render('Subtree_Sampling.Rmd')"  

Rscript -e "rmarkdown::render('Subtree_Sampling.Rmd')" fastadir="in.fasta", outdir="out.trees", pwd_rank_cutff=0.47

#### Input and the default values

1. fastadir ("in.fasta" by default): directory where fasta sequence files are located. The name of each fasta file consists of subtype + ".txt". ie, 01_AE.txt, B.txt. The name of each sequence in a fasta file consists of sequenceid (in this case, accessionid), country (in two letter ISO code), collection year (YY), subtype delimited by "."

2. outdir ("out.trees" by default): directory for saving output files including .tre, .pdf, and .txt of sampled sequences

3. pwd_rank_cutff(0.47 by default): a node of which a local MPWD (MPWD of the node's children) is less than the global MPWD at the rank of this cutoff defined as a cluster (subtree)

#### Output

1. Subtree_Sampling.html

2. results.summary.txt: lists number of clusters, number of sequences sampled so on.

Files generated in outdir for each subtype:

3. a tree of the entire sequences (.tre) (.pdf)

4. a tree of the entire sequences with cluster ids defined by a cutoff (.tre)

5. a fasta sequence file containing sampled sequences (.txt)


#### Packages required

"pandoc" should be installed:

https://pandoc.org/installing.html

R packages:

rmarkdown, knitr, ape, igraph, reshape2 and phangorn

```{r set-options,include=FALSE}
library("knitr")
knitr::opts_chunk$set(echo = TRUE)
options(width = 1800)

library(ape, quietly=TRUE)
library(igraph, quietly=TRUE)
library(reshape2, quietly=TRUE)
library(phangorn, quietly=TRUE)


## ------------Functions------------------

## Given a tree and a node (node index), 
## return a list of indexes of the node's leaves (a recursive function)
node.tips<-function (tree, node){
  n <- Ntip(tree)    # number of leaves, also the last index of the leaves
  if (node <= n){
      # Because node numbering of a tree 
      # starts with all the leaves first and the internal nodes 
      # starting from the root, it is a leave
      # return the index
      node
  }
  else{
      # for an internal node
      l<- numeric()
      # get all edges connecting from the node to children nodes
      d<- tree$edge[which(tree$edge[,1]==node),2]
       for(j in d){
          if(j <= n){
            # a leave, add it to a list
            l <- c(l, j)
          }
          else{
            # an internal node, goes down to another recursion
            l<-c(l, node.tips(tree,j))
          }
      }
       l
   }
}

## Given a node, tree, and distance matrix,
## return median pairwise distance (MPWD) of the node's leaves
get.node.leaf.MPWD <- function(node,tree,distmat){
  ## get indexes of all tips under a node (internal node)
  nlist <- node.tips(tree,node)  
  foo <- distmat[nlist,nlist]
  return(median(foo[upper.tri(foo,diag=FALSE)]))
}


## Given a tree and a distance matrix of leave nodes
## return a vector containing the median pairwise distance 
## of the subtree under each internal node
mpwd.clusttree <- function(tree, distmatrix){
  distmat0 <- as.matrix(distmatrix)
  # get pairwise distances between leaves (sequences)
  distmat <- distmat0[tree$tip.label,tree$tip.label]

  ntips<- Ntip(tree)                  # number of tips
  nint <- Nnode(tree)                 # number of nodes
  
  # node numbering of a tree starts with all the leaves first 
  # and the internal nodes starting from the root
  # so the index of a first internal node is ntips+1 
  # the index of a last internal node is ntips+nint
  # therefore, indexes of all internal nodes are
  # ntips+1:ntips+nint
   return(sapply((ntips+1):(ntips+nint),get.node.leaf.MPWD,tree,distmat))
}


## Given a tree, a threshold or maximum MPWD of a subtree and a distance matrix 
## return a vector indicating, for each node, which cluster the node belongs to
## These functions were borrowed from 
## https://www.r-bloggers.com/finding-meaningful-clusters-in-phylogenetic-trees-or-other-hierarchical-clusterings/
## Instead of using patiristic distances, the modified codes here use pairwise distances
clustering <- function (tree, thresh,distmatrix){
  ## a vector of MPWD of leaves of all internal nodes
  distvec <- mpwd.clusttree(tree, distmatrix) 

  ## set up clustering
  ntips<-Ntip(tree)
  nint <- Nnode(tree)
  cnum <- 0                                  ## cluster id
  assign <- rep(0,ntips+nint)                ## cluster assignment
  
  ## tree in igraph form (we need to use igraph for depth first search of the tree)
  igraph.tree <- graph.edgelist(tree$edge) 
  
  ## to retrieve nodes in the DFS order starting from the root
  dfs <- graph.dfs(igraph.tree,root=ntips+1,neimode='out',order=TRUE,dist=TRUE)                  
                   
  ## traverse the tree in depth first order
  for(i in 1:length(dfs$order)){
    node <- dfs$order[i]
    ## skip leaves
    if(node <= ntips){ next }
    
    ## If the node's subtree is below the threshold, mark it and
    ## its subtree as members of a new cluster
    ## node-ntips --> node numbering of a tree starts with all the leaves first 
    ## and the internal nodes starting from the root
    ## but the distvec contains only MPPD to leaves of internal nodes
    if(distvec[node-ntips]<=thresh && assign[node] == 0){
      cnum <- cnum+1
      subtree <- graph.dfs(igraph.tree,node,
                           neimode='out',unreachable=FALSE)$order
      subtree <- subtree[!is.na(subtree)]
      assign[subtree] <- cnum
    }
  }
  clusteringResults <- list(membership=assign, allcsize=table(assign),
              membershipJustLeaves=assign[1:ntips], leafclustsize=table(assign[1:ntips]),
              ntips=ntips,threshold=thresh)

  return(clusteringResults)
}

## Number of clusters = number of leaves with clid 0 (singleton) +
##                      number of unique clids
num_clusters <- function (dat){
  # a leave with clid =0 is a sigleton; count each one as a cluster
  return (nrow(dat[dat$clid == 0, ]) + length(unique(dat$clid[dat$clid != 0])))
}


## For each cluster, select one per country
choose_seq_one_per_country <- function(dat){
  ## include all singletons
  dat1 <- dat[dat$clid == 0, ]
  clidlist <- unique(dat$clid[dat$clid != 0])
  
  for(clid in clidlist){
    regions <- unique(dat[dat$clid == clid, "ISOCode"])
    for(r in regions){
       dat1 <- rbind(dat1, dat[dat$clid == clid & dat$ISOCode == r, ][1,])
    }
  }
  return (dat1)
}

## Distribution of pairwise distances 
pairwise_dist_distribution <- function (distmatrix){
  distmatrix <- as.matrix(distmatrix)
  dtable <- melt(distmatrix)[melt(upper.tri(distmatrix))$value,]
  names(dtable) <- c("s1", "s2", "distance")
  numpairs <- nrow(dtable)
  mycaption <- paste(s, ": ", numseqs, " sequences, ", numpairs, " pairs", sep="")
  hist(dtable$distance, main=mycaption, xlab="pairwise distance", ylab="number of sequence pairs")
  qu<-quantile(dtable$distance, c(.25, .45, .47, .48, .49, .5, .75))
  print(qu)
}

## Save a tree in .tre abd .pdf
save_tree<-function(tree, s){
  ## Save it in Newick format
  treefile <- paste(outdir, "/", s, ".tre", sep="")
  write.tree(tree, file=treefile)
  ## Save it as a pdf
  pdffile <- paste(outdir, "/", s, ".pdf", sep="")
  pdf(file=pdffile)
  plot(tree, cex=0.4, type="phylogram")
  add.scale.bar(cex = 0.7, font = 2, col = "red", lcol="red")
  dev.off()
}

##' commandArgs parsing
##' 
##' return a named list of command line arguments
##'
##' Usage:
##' call the R script thus
##'   ./myfile.R --args myarg=something
##' or
##'   R CMD BATCH --args myarg=something myfile.R
##'
##' Then in R do
##'   myargs <- getArgs()
##' and myargs will be a named list
##' > str(myargs)
##' List of 2
##' $ file : chr "myfile.R"
##' $ myarg: chr "something"
##'
##' @title getArgs
##' @param verbose print verbage to screen 
##' @param defaults a named list of defaults, optional
##' @return a named list
##' @author Chris Wallace
getArgs <- function(verbose=FALSE, defaults=NULL) {
  myargs <- gsub("^--","",commandArgs(TRUE))
  setopts <- !grepl("=",myargs)
  if(any(setopts))
    myargs[setopts] <- paste(myargs[setopts],"=notset",sep="")
  myargs.list <- strsplit(myargs,"=")
  myargs <- lapply(myargs.list,"[[",2 )
  names(myargs) <- lapply(myargs.list, "[[", 1)

  ## logicals
  if(any(setopts))
    myargs[setopts] <- TRUE

  ## defaults
  if(!is.null(defaults)) {
    defs.needed <- setdiff(names(defaults), names(myargs))
    if(length(defs.needed)) {
      myargs[ defs.needed ] <- defaults[ defs.needed ]
    }
  }

  ## verbage
  if(verbose) {
    cat("read",length(myargs),"named args:\n")
    print(myargs)
  }
  myargs
}
```

## Analysis

#### Processing command-line input

```{r echo=FALSE}
## input and default values
fastadir = "in.fasta"
outdir = "out.trees"
pwd_rank_cutoff <- 0.47
summaryfile <- "results.summary.txt"

## clustering is performed only for the trees with median PWD (MPWD) >= "minPWD". 
minPWD = 0.015

## read in from command-line
args <- getArgs(defaults=list(fastadir=fastadir, outdir=outdir, pwd_rank_cutoff=pwd_rank_cutoff))
fastadir <- args$fastadir
outdir <- args$outdir
pwd_rank_cutoff <- as.numeric(args$pwd_rank_cutoff)

cat("\tfastadir: ", fastadir, "\n")
cat("\toutdir: ", outdir, "\n")
cat("\tpwd_rank_cutoff: ", pwd_rank_cutoff, "\n")

if(!file.exists(fastadir)){
    cat("Error!! fastadir ", fastadir, " doesn't exist")
    knitr::knit_exit()
}

## if the outdir doesn't exists, create one
dir.create(file.path(".", outdir), showWarnings = FALSE)
```


#### Processing input sequence files

```{r echo=FALSE}
fastafiles <- list.files(fastadir)
numfiles <- length(fastafiles)
subtypes <- sub(".txt","", fastafiles)
numsubtypes <- length(subtypes)
cat("No. fasta files: ", numfiles)
cat("No. subtypes: ", numsubtypes)

# For saving overall clustering results for each subtype/CRF
results <- data.frame(Subtype = character(), NumSeqs = numeric(),
                      NumCountries = numeric(), PWDCutOff = numeric(),
                      NumCluster = numeric(), NumSelected = numeric(), 
                      stringsAsFactors =FALSE)

i<-0
for(fastafile in fastafiles){
    s <- sub(".txt","", fastafile)
    cat("Subtype: ", s,"\n")
    
    fastafilepath = paste(fastadir, "/", fastafile, sep="") 
    seqdata <- read.dna(file=fastafilepath, format = "fasta", as.matrix=TRUE)
    numseqs <- nrow(seqdata)
    cat("no. sequences: ", numseqs,"\n")
    # Can't create a tree for subtype/CRF with <=3 sequences 
    if(numseqs <=3) {next}
    
    # Convert sequence names to a data.frame
    seqnames <- rownames(seqdata)
    d <- as.data.frame(matrix(unlist(strsplit(seqnames, "[.]")), ncol=4,byrow=TRUE))
    colnames(d) <- c("Accession", "ISOCode", "Year", "Subtype")
    
    numcountries <- length(unique(d$ISOCode))
    cat("no. countries: ", numcountries,"\n")
    
    # Create a pairwise distance (PWD) matrix using TN93 evloution model
    # to get the distribution of PWDs 
    distmatrix <- dist.dna(seqdata, model="TN93", pairwise.deletion = TRUE)
    pairwise_dist_distribution(distmatrix)
    
    # Create a NJ tree 
    tree <- NJ(distmatrix)
    # Root at the midpoint of the longest path between any two tips
    tree <- midpoint(tree)      
    # For some branches, the length can be negative.
    # Change to zero for branches with a negative length (as recommended) 
    tree$edge.length[tree$edge.length < 0] <- 0
    # Plot the tree  
    maintext <- paste("subtype ",s, ": ", numseqs, " sequences", sep="")
    plot(tree, cex=0.7, main=maintext)
    #nodelabels()
    add.scale.bar(cex = 0.7, font = 2, col = "red", lcol="red")
    
    save_tree(tree, s)

    if(median(distmatrix) < minPWD){ next}
    
    # The MPWD at the rank of "pwd_rank_cutoff"
    cluster_pwd_cutoff <- quantile(distmatrix, c(pwd_rank_cutoff))
    cat("PWD at the rank of ", pwd_rank_cutoff,": ", cluster_pwd_cutoff,"\n")
    
    # Identify subtrees (clusters) of sequences of which their median PWD < cluster_pwd_cutoff 
    # and assign a cluster id number to each sequence belonging to a same cluster
    subtrees <- clustering(tree,cluster_pwd_cutoff,distmatrix)
  
    ## Add cluster ids to at the begining of sequence labels in the tree
    treeWithClusterIDs <- tree
    treeWithClusterIDs$tip.label <- subtrees$membershipJustLeaves
    plot(treeWithClusterIDs, cex=0.6, main=paste("threshold",  cluster_pwd_cutoff, sep=' '))
    add.scale.bar(cex = 0.7, font = 2, col = "red", lcol="red")
    treefile <- paste(outdir, "/", s, "_", pwd_rank_cutoff, ".tre", sep="")
    write.tree(treeWithClusterIDs, file=treefile)

    d$clid <- subtrees$membershipJustLeaves
    d$seqid <- tree$tip.label
    
    ## Choose one per country per cluster
    numclusters <- num_clusters(d)
    cat("no. clusters: ", numclusters, "\n")
    d0<- choose_seq_one_per_country(d)
    numchosen <- nrow(d0)
    cat("no. sequences sampled: ", numchosen,"\n")
  
    ## Save the sampled sequences in a fasta file
    fname <- paste(outdir, "/", s, "_", pwd_rank_cutoff, ".txt", sep="")
    write.dna(seqdata[as.character(d0$seqid),], file=fname, format='fasta', colsep='')
    
    i <- i+1
    results[i,] <- c(s,numseqs,numcountries,cluster_pwd_cutoff,numclusters, numchosen)
    cat("\n\n\n\n")
}

results <- results[order(results$Subtype, decreasing = FALSE), ]
write.table(results, file=summaryfile, sep="\t", row.names=FALSE, quote=FALSE)

```


